---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# StatCompLab

<!-- badges: start -->
<!-- badges: end -->

This package collects workshop materials for Statistical Computing (MATH10093)
at the University of Edinburgh 2021/22.

The tutorial documents are browsable online at [https://finnlindgren.github.io/StatCompLab/](https://finnlindgren.github.io/StatCompLab/)

Contact Finn Lindgren, [finn.lindgren@ed.ac.uk](mailto:finn.lindgren@ed.ac.uk)
for information.

The package version number system is `YearOfStudy`.`StudyWeek`.`MinorUpdate`

## Installation

You can install (and later upgrade) the package from
[GitHub](https://github.com/) with:

``` r
# If devtools isn't installed, first run install.packages("devtools")
devtools::install_github("finnlindgren/StatCompLab", build_vignettes = TRUE)
```
## Example

```{r example,eval=FALSE}
library(StatCompLab)
vignette(package = "StatCompLab") # List available vignettes
vignette("Tutorial01", "StatCompLab") # View a specific vignette
library(learnr)
available_tutorials("StatCompLab") # List available tutorials
run_tutorial("Tutorial01", "StatCompLab") # Run a specific tutorial
```
