---
title: "Project 01 Hints"
author: "Finn Lindgren"
date: "2022-02-11"
output:
  rmarkdown::html_vignette:
    toc: true
  pdf_document:
    toc: true
vignette: |
  %\VignetteIndexEntry{Project 01 Hints}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Avoiding long running times

## The problem

Collecting values into a data.frame inside a loop can be slow.

A common issue when combining results computed in loops is that using `rbind` in each step of the loop to add one or a few rows to a large vector or data.frame can be very slow; instead of linear computational cost in the size of the output, the cost can become quadratic; it needs to reallocate memory and copy the previous results, for each new version of the result object.

## The solution

The key is to avoid reallocating memory by instead pre-allocating the whole result object before the loop. However there are some subtleties to this, and the examples blow show the differences between approaches.

A common solution is to work with simpler intermediate data structures within the loop, and then collect the results at the end.
Indexing into simple vectors is much cheaper than reallocating and copying memory.

Slow:
```{r}
slow_fun <- function(N) {
  result <- data.frame()
  for (loop in seq_len(N)) {
    result <- rbind(
      result,
      data.frame(A = cos(loop), B = sin(loop))
    )
  }
  result
}
```

Better, but even data.frame indexing assignments causes some memory reallocation:
```{r}
better_fun <- function(N) {
  result <- data.frame(A = numeric(N), B = numeric(N))
  for (loop in seq_len(N)) {
    result[loop, ] <- c(cos(loop), sin(loop))
  }
  result
}
```

Fast; several orders of magnitude faster (see timings below) by avoiding memory reallocation inside the loop:
```{r}
fast_fun <- function(N) {
  result_A <- numeric(N)
  result_B <- numeric(N)
  for (loop in seq_len(N)) {
    result_A[loop] <- cos(loop)
    result_B[loop] <- sin(loop)
  }
  data.frame(
    A = result_A,
    B = result_B
  )
}
```

Not as fast, but a bit more compact; using matrix indexing doesn't need memory reallocation, but is a bit slower than vector indexing. A potential drawback is that his method only works when all the columns in the output data.frame should be of the same type (numeric), but
for cases where it works the code is both simple, readable, and generally quick to run:
```{r}
ok_fun <- function(N) {
  result <- matrix(0, N, 2)
  for (loop in seq_len(N)) {
    result[loop, ] <- c(cos(loop), sin(loop))
  }
  colnames(result) <- c("A", "B")
  as.data.frame(result)
}
```

Benchmark timing comparisons:
```{r}
N <- 10000
bench::mark(
  slow = slow_fun(N),
  better = better_fun(N),
  fast = fast_fun(N),
  ok = ok_fun(N)
)
```

# Dynamically generated column names

When the size of a matrix or data.frame depends on a parameter, or there are multiple columns with similar names, it can be convenient to dynamically generate column names. The `paste()` function can be used for this.

For example, if we want a data.frame of size $n\times m$, with column names `Name1`, `Name2`, etc, we can use
```{r, eval=TRUE, echo=TRUE}
df <- matrix(runif(15), 5, 3) # Create a matrix with some values in it
colnames(df) <- paste("Name", seq_len(ncol(df)), sep = "")
df <- as.data.frame(df)
```
The resulting object looks like this:
```{r, eval=TRUE, echo=TRUE}
df
```

# Avoiding unnecessary for-loops

Since most operations and functions in R are vectorised, the code can often be
written in a clear way by avoiding unnecessary for-loops. As a simple example,
say that we want to compute $\sum_{k=1}^{100} \cos(k)$.

A for-loop solution of the style
that would be a good solution in a high performance low-level language like C
might look like this:
```{r, eval=FALSE}
result <- 0
for (k in seq_len(100)) {
  result <- result + cos(k)
}
```
By taking advantage of vectorised calling of `cos()`, and the function `sum()`,
we can reformulate into a shorter and more clear vectorised solution:
```{r, eval=FALSE}
result <- sum(cos(seq_len(100)))
```

Other commonly used functions involving vectors of `TRUE` and `FALSE` are `all()`
and `any()`. For example,
```{r}
vec <- c(TRUE, TRUE, FALSE, TRUE)
all(vec)
any(vec)
sum(vec) # FALSE == 0, TRUE == 1
```

Both `sum`, `any`, and `all` also operate across all elements of a matrix.
For row-wise and column-wise sums, see `rowSums()`and `colSums()`.

# LaTeX equations

## Aligning equations in multi-step derivations

When typesetting multi-step derivations, one should align the equations. In
LaTeX, this can be done with the `align` or `align*` environments.
In some case, that might not work in RMarkdown, but one can then use the
`aligned` environment instead.

### Example 1

```
\begin{align*}
\frac{\partial}{\partial x}f(x, y) &= \frac{\partial}{\partial x}\exp(2x+3y) \\
&= 2\exp(2x+3y)
\end{align*}
```
Result:

\begin{align*}
\frac{\partial}{\partial x}f(x, y) &= \frac{\partial}{\partial x}\exp(2x+3y) \\
&= 2\exp(2x+3y)
\end{align*}

### Example 2

```
$$
\begin{aligned}
\frac{\partial}{\partial x}f(x, y) &= \frac{\partial}{\partial x}\exp(2x+3y) \\
&= 2\exp(2x+3y)
\end{aligned}
$$
```
Result:

$$
\begin{aligned}
\frac{\partial}{\partial x}f(x, y) &= \frac{\partial}{\partial x}\exp(2x+3y) \\
&= 2\exp(2x+3y)
\end{aligned}
$$
