---
title: "Project 02 Hints"
author: "Finn Lindgren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Project 02 Hints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Extracting information from linear model objects

When we've estimated a linear model with `lm()`, we often want to extract information
to be used in presentation of results (tables and figures) or in further calculations.
You've already used the `predict()` function for calculating predictions and
prediction uncertainties. In Project 2, you also need to access information
about the estimated parameters and their uncertainty.

Let's say we've estimated a simple model:
```{r}
fit <- lm(y ~ 1 + x, data = data.frame(x = rnorm(10), y = rnorm(10)))
```

## Basic option

The coefficient estimates can be extracted with
```{r}
coef(fit)
coef(fit)["x"]
```

The estimated covariance matrix of the coefficient estimates can be extracted with
```{r}
vcov(fit)
```
To get Standard Error for a coefficient, extract the diagonal element and take
the square root.
```{r}
sqrt(vcov(fit)["x", "x"])
```

The coefficient information can be combined with standard errors to construct approximate
confidence intervals, which in the project can be plotted with `geom_ribbon()`.

The coefficient estimates and uncertainty information can also be extracted with
```{r}
summary(fit)$coefficients
```
To extract specific elements:
```{r}
coefs <- summary(fit)$coefficients
coefs["x", "Estimate"]
coefs["x", "Std. Error"]
```

## Tidy option

The `broom` package can provide the same information in a similar way, but for
this particular type of problem it doesn't really add any value over the basic
approach:
```{r}
suppressPackageStartupMessages(library(tidyverse))
library(broom)
tidy(fit)
tidy(fit) %>%
  filter(term == "x") %>%
  pull(std.error)
```
