---
title: "Project 1 hints"
author: "Finn Lindgren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Part 1

## Avoiding long running times

### The problem

Collecting the confidence intervals into a data.frame inside a loop can be slow.

A common issue when combining results computed in loops is that using `rbind` in each step of the loop to add one or a few rows to a large vector or data.frame can be very slow; instead of linear computational cost in the size of the output, the cost can become quadratic; it needs to reallocate memory and copy the previous results, for each new version of the result object.

### The solution

The key is to avoid reallocating memory by instead pre-allocating the whole result object before the loop. However there are some subtleties to this, and the examples blow show the differences between approaches.

A common solution is to work with simpler intermediate data structures within the loop, and then collect the results at the end.
Indexing into simple vectors is much cheaper than reallocating and copying memory.

Slow:
```{r}
slow_fun <- function(N) {
  result <- data.frame()
  for (loop in seq_len(N)) {
    result <- rbind(
      result,
      data.frame(A = cos(loop), B = sin(loop))
    )
  }
  result
}
```

Better, but even data.frame indexing assignments causes some memory reallocation:
```{r}
better_fun <- function(N) {
  result <- data.frame(A = numeric(N), B = numeric(N))
  for (loop in seq_len(N)) {
    result[loop, ] <- c(cos(loop), sin(loop))
  }
  result
}
```

Fast; several orders of magnitude faster (see timings below) by avoiding memory reallocation inside the loop:
```{r}
fast_fun <- function(N) {
  result_A <- numeric(N)
  result_B <- numeric(N)
  for (loop in seq_len(N)) {
    result_A[loop] <- cos(loop)
    result_B[loop] <- sin(loop)
  }
  data.frame(
    A = result_A,
    B = result_B
  )
}
```

Not as fast, but a bit more compact; using matrix indexing doesn't need memory reallocation, but is a bit slower than vector indexing. A potential drawback is that his method only works when all the columns in the output data.frame should be of the same type (numeric).
```{r}
ok_fun <- function(N) {
  result <- matrix(0, N, 2)
  for (loop in seq_len(N)) {
    result[loop, ] <- c(cos(loop), sin(loop))
  }
  colnames(result) <- c("A", "B")
  as.data.frame(result)
}
```

Benchmark timing comparisons:

```{r}
N <- 1000
bench::mark(
  slow = slow_fun(N),
  better = better_fun(N),
  fast = fast_fun(N),
  ok = ok_fun(N)
)
```

```{r}
N <- 10000
bench::mark(
  slow = slow_fun(N),
  better = better_fun(N),
  fast = fast_fun(N),
  ok = ok_fun(N)
)
```
